---
title: "Intro to R for Capstone 2020"
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---


### What is R? What is R Studio?

R refers to both the programming language and software used for statistical computing.  

R Studio is a graphical user interface (GUI) that makes it much easier to interact with the R software. It adds handy drop-down menus, visualizations, and other features that make your life easier. R does the work, R Studio makes it pretty. 


### Why R?

1. **It's Free!:** Obviously a benefit since other statistical programs require subscriptions in the range of thousands of dollars. It's also open-source which means anyone can inspect is and report bugs, reducing the chance of mistakes. 

2. **Its reproducible!:** R involves a lot more typing than pointing and clicking. While this means it might be a steeper learning curve, it makes it way easier to repeat an analysis. Instead of having to remember what all you clicked on and in what order, you can just re-run code you typed and saved. This means someone else (or your future self) can easily repeat your analysis and obtain the same results. 

3. **Lots of people use it!:** and many of them are willing to help you out. There are tons of online (and free) resources for learning R.

4. **It can do almost anything!:** R works on data of all shapes and sizes and produces high quality (even interactive) graphics and figures. There are tens of thousands of available packages that extend the capabilities of R. Nearly anything you can think of, there's an R package that can do it. 


### Finding your way around R Studio

After opening R studio you will see 4 panes (maybe 3, meaning one is hidden and that's fine). These four panes are as follows and their locations are given based upon the defaults of R Studio. You can change this (or may have already done so) if you prefer an alternate set up. 

1. Source (Top Left). This pane contains the scripts, markdown files, or data you are working on currently. You can have more than one open at a time.
      
2. Console (Bottom Left). This pane contains a history of the commands you have ran. You can also write code directly in this pane, though not always recommended. Any code you write in the console will not save, so we recommend doing all of your coding in a script (top left pane).
      
3. Environment/History/Connections (Top Right). The environment contains data which R can access at this given moment. The history tab contains a history of the code you have ran. The connections tab allows you to access outside data sources.
      
4. The final pane (Bottom Right) contains several tabs. In this pane, you can view plots, search for files on your computer, install packages (we will get to this in a minute), and even find documentation on packages or functions. This pane is your friend. Become comfortable with it, and you will use it often.

Personalizing your Studio experience:
Under the "Tools" menu (at the top of the screen) you can choose "Global Options" which allows you to change various things including the layout of the panes described above ("Pane Layout") and the background color/theme ("Appearance").


### R scripts & R Markdown files

One of the major  benefits of using a program like R is the ability to write code once and "reuse" it over and over again with slight modifications. This allows you to run an analysis for a collaborator (or yourself) and save it until next year when they ask you to do the same thing again. This is also essential for reproducibility. You can type code directly into the console, but this would only save your work until you exit the program. In order to save your work for future analyses, programming into a notebook or script is highly recommended. For example, this R script is written in a special kind of notebook known as markdown. This format, which is different from an R script, allows programmers to include text, figures, headings, and code all in the same place which makes for easy distribution. In addition, this notebook can be rendered ("knitted") in a .html format (or even a PDF, Word, or PowerPoint file!) so anyone, even without R software, can access it. This is ideal for tutorials so we both run and explain our code and view and share it in a nice-looking format later. 
 
Obviously this part of this markdown file you are reading is text and the box below is a "chunk" of code. You can create a code chunk by pressing **Cmd + Option + I** (Mac) or **Ctrl + Alt + I** (Windows) or typing ```{r} then ```. When you run code in a markdown file, the result (including plots) will appear directly below the code chunk. You can run the below chunk of code by pressing **CTRL + ENTER** (**OPTION + ENTER** and **COMMAND + ENTER** also work on a Mac) or by clicking the green arrow in the upper right corner of the code chunk.

```{r}
print ("This is code")
#this is a comment
```

R Scripts are also quite useful for saving code. These files are a bit more "simple" than R markdown files in that they contain only code and any notes/comments must be "commented out" using a # symbol. Another difference is where the result appears. When using an R script, figures generally appear in the plots pane of R studio and other calculations appear in the console. For code for you actual analyses, I would recommend a R Script over a markdown file, but it's really a personal preference. 
Feel free to open a new script and/or markdown file to see these differences. 


### R Projects and Working Directories

Your working environment and working directory are extremely important when you want to read in data or save an analysis. Imagine you are in your office; this is your working directory. You can write on a piece of paper and save it on your desk. You can also find notes which you left yourself. You are able to easily access anything in your office. Your neighbors office, while located in the same hallway, requires you to back out of your office and then enter hers. This analogy holds true for finding files on your computer as well. Your working environment will allow you to access files quickly and to write copies of your analysis or environment easily. Like you can walk into your co-workers office, you can change your working directory to access a different set of files. 

I am a big fan of using "R Projects" this is especially helpful when working on multiple things in R (various classes and research projects). An R project is housed within a directory and that directory is automatically set as your working directory. When you save your project, you essentially save everything in each pane of R Studio: this includes the scripts you are working with as well as anything in your working environment. This allows you to close an R session with multiple files and data sets and open it up the next day exactly how you left it. My only caution about this method is that this can make you a bit lazy when it comes to defining your working directory and well as keeping things neat and organized. 

I would recommend creating a folder (aka directory) on your computer dedicated to this class and create an R Project within that folder. You can do this by choosing "New Project" under the "File" menu and navigating to the desired folder. **Make sure this file is saved in that directory and any associated data should also be saved there.** If you run the code below (by clicking the green arrow) it should return this same file path. 

```{r}
#R will not interpret this line due to the # at the begining of the line, this is called commenting
#Lets see where we are right now (our current working directory)
getwd()
```
 
If you ever need to change your working directory (analogous to walking down the hall to your neighbor's office), you can use the below code. If you run this code as is, you will get an error message because the file path I wrote does not exist on your computer (I hope...). I don't recommend re-setting your working directory unless you need to as this can cause problems and confusion.

```{r eval=FALSE, include=FALSE}
#Now lets say we want to change our working directory to your desktop folder on your machine. This will only change the working directory in this chunk of code. Once this chunk finishes running your working directory will revert to whatever it was prior to running this chunk.
setwd("/Users/Ella/Documents/College Crap/Research/Data Analysis Resources/Capstone2020")
``` 


### About this Tutorial

This tutorial is a mixture of information, provided example code and "challenges" like the one above in which you will write your own code (which may be required for future steps...). Don't be scared to ask for help on any of these!! The overarching goal of this tutorial is for you to create your own resource to refer to throughout this course and any future coding you might do. This R Markdown file can be "knitted" to an html file, just like many of the tutorials and resources you might find online. Go ahead and click on "preview" at the top of this panel in R Studio to get a glimpse of what your final product will look like. You will also notice an html file with the name will appear in the same directory as this file. All you need to view this file is an internet browser. As you work through, run the code provided, and add some of your own code, this will fill in the document. I highly encourage you to add your own notes either in the non-code portions or as commented code. This is your resource, so personalize it and make it whatever you feel will help you the most!

You do not need to memorize all of the information here! The key is becoming familiar with R and knowing where you can find the code that will accomplish what you need to do whether it's referencing this tutorial, using the help pages built into R, or a well-worded Google search.


### Lets get started!!!

### R as a calculator

The most basic function of R is as a normal calculator. Below are some examples of basic operations you can perform in R (and really any thing you can think of, you can do it in R). You can run each of these calculations individually by placing your cursor on the line you want and pressing **CTRL + ENTER** (**OPTION + ENTER** and **COMMAND + ENTER** also work on a Mac). Alternatively, you can run all of them at the same time by clicking the green arrow in the upper right corner of the code chunk. The same goes for any other chunk of code in here. 

```{r}
2+7
9-1
2*12
5/2
6^2
```


### Creating Objects in R

When you start doing more complex calculations that use much more computer power, you won't want to run the function every time you need it. Instead, you can store the values and name them using the "assignment operator" (<-). Think of this as an arrow assigning anything on the right to the name on the left. Below is an example. Notice that when you run this code, "x" appears in your global environment (the upper right panel in my R Studio).

```{r}
x <- 1 + 1
```

Notice that this does not show us the value of x. There are a few ways we can tell R to print the value. 

(Also remember you can (and definitely should) use # to insert comments into your code as demonstrated below)

```{r}
(x <- 1 + 1)  #parentheses will tell R to create an object and print the value all in one line
x #after creating our object "x" we can tell R to print the value by typing the name
```


#### *Challenge #x:*
##### *We can now perform further operations on the objects we create. Can you create an object named "y" that is 3 times the value of x?*

```{r}
y <- 3*x
```


### Functions

Anything followed by a set of parentheses in R is a function. R comes with many predefined functions and way more are available through packages. When we install and loading packages, we are essentially just defining functions. The below example is a simple function that will take the square root of a number.

```{r}
sqrt (69)
```

The above function only takes one argument (the number you want to take the square root of) but many functions take multiple arguments. Another simple example is the round() function which takes two arguments: the number you want to round (x), and the number of resulting decimal places you would like (digits). These can be defined explicitly or implicitly in a specified order as shown below. (Also notice that pi is pre-defined in R).

```{r}
round (x=pi, digits=2) #defined explicitly
round (pi, 2) #defined implicitly, ORDER MATTERS when you define like this
round (digits=2, x=pi) #order does not matter when you define explicitly
```

You can also nest or "pipe" functions in which you send the result of one function into another. For example, to round the square root of a number, you could do the following:

```{r}
round (sqrt (69), digits=2)
```


#### Installing Packages

Packages are basically a collection of functions that aren't included in base R, which allows us to access a lot more functionality of R without having to write all of the code from scratch ourselves. To access a package you need to do two things: install it and load it. You only need to install a package once on your computer (unless you need to re-download or update R) but you need to load the package each time you open R and wish to use it. Run the code below to install the "lubridate" package. We will use this later in the tutorial.

```{r}
#this line downloads the package to your computer
install.packages("lubridate")
#this line loads the package into your current session
library(lubridate)

#this lets you see what packages you have loaded at the moment
(.packages())
```


##### Getting Help

I've said it before but I will say it again: you do not need to memorize any of the functions in R, you just need to know where to find the necessary information about them. One easy way to do this is the help page in R. If you remember the name of a function but don't remember what it does or the arguments it takes, or see a function you are not familiar with, simply type the name of the function with a question mark in front of it to access the help page for the function. Run the code below to learn about the "plot()" function in R (one that we will use quite a bit...). What arguments does this function require?

Another great resource is Google! It is 100% OK to Google things like "scatter plots in R" or "plot() function in R". 

```{r}
?plot
```


### Vectors, Matrices, and Data Types

Up until now, we have saved single values when we create objects, but we can do much more. Vectors are one of the most common data types in R and are composed of a series of values. To create a vector we use the c() function. 

```{r}
v <- c(1, 5, 2, 8, 3)
w <- 1:5 #this is a shortcut for "w <- c(1, 2, 3, 4, 5)
```

Vectors can also be character values (aka letters, words, or a combination of letters and numbers). Note that the quotes are necessary, it tells R to treat these words as something called character strings rather than looking for an object named "blackberries" that does not exist.

```{r}
fruit <- c("blackberries", "raspberries", "peaches", "plums")
```


#### Functions on Vectors

We can perform functions on our vectors. Some functions act on each individual element of the vector:

```{r}
v + 2
v*8
v + w #notice that these vectors must be the same length to do this...see how to find the length of a vector below
```

...while others act on the vector as a whole.

```{r}
sum (v)
length (v)
```


#### *Challenge #x:*
##### *Can you find the mean of our vector v?*

```{r}
mean (v)
```


#### Specifying elements in a vector

You can refer to specific numbers within your vector using "indexing" with square brackets. For exampling, if you wanted to know the 3rd number in the vector v, you would run the following code:

```{r}
v[3]
```

#### *Challenge #x:*
##### *Use the above indexing to find the 4th fruit in our "fruit" vector?*

```{r}
fruit [4]
```


#### Hierarchy of Data Types

As shown above, vectors can be made of numbers *or* characters. However, a single vector must contain all the same data type. There are several other types of data R recognizes and there is a hierarchy of how R will coerce data (shown below). If a vector contains elements of different data types, R will treat it as the lowest type on the hierarchy. For example, if you mix letters and numbers, R will treat every element as a character value.

1. Logical (TRUE or FALSE (or just T or F))

2. Integer (1, 2, 3...)

3. Numerical (any number)

4. Complex Numbers (numbers with imaginary components...remember i?)

5. Character or factor (letters)

You can use the str () or class() functions to determine the type of vector you are working with. For example

```{r}
str (v)
```


##### *Challenge #x:*
##### *What types of data are each of the vectors defined below? Make sure you understand and could explain why.*

```{r}
a <- 1:4
b <- c("1", "2", "3", "4")
c <- c (T, F, FALSE, TRUE, NA)
d <- c ("a", "b", "c", 4)
e <- c("A", "B", "C", "D", "E", F)
```

```{r}
str (a) #integer
str (b) #character
str (c) #logical
str (d) #character
str (e) #character
```


#### Matrices

Vectors are data in 1 dimension, matrices are data in 2 dimensions. You can even go farther with arrays which are data in 3 or more dimensions (but that hurts my brain to think too much about that...). We won't spend too much time with matrices but it's good to know they exist and how they work. Here are a few examples. Notice how each element is indexed [row, column].

```{r}
matrix (1:9, nrow=3, byrow=TRUE)
matrix (1:9, nrow=3, byrow=FALSE)
```


### Data Frames!!!

Think of data frames as the R version of excel spreadsheets. They are essentially a collection of vectors that are all the same length. As such, we can create data frames in R from vectors.

```{r}
#first we create our vectors
names <- c("Sally", "Bob", "Joe", "Susan")
height_in <- c(66, 70, 72, 69)
age <- c(30, 45, 23, 62)

#now we combine them into a data frame
data <- data.frame(names=names, height_in=height_in, age=age)
#the reason we say names=names is that we want a column called "names" 
#and the data for it can be found in the object called "names".
```

Now lets take a look at our data frame. We can do this several ways. Try running the below code and clicking the spreadsheet icon near the "data" object in your environment (top right pane).

Ask yourself: Does the data look like what we might expect or want it to? What type of data is present in each column? 

```{r}
data #this prints the data frame straight into the R Markdown
View (data) #this opens a tab to view your data frame (the same as clicking on it)
str (data) #this tells you the types of data present in each column
```


#### Subsetting Data

The first way to subset data is by numerically specifying rows and columns in the format data [row, column], similar to what we did earlier with vectors and matrices. For example:

```{r}
data[1,] #first row, all columns
data[,1] #first column, all rows
data[1,2] #first row, second column
```

You can (thankfully) also specify columns by name sing a dollar sign:

```{r}
data$names
```

We can also subset using logical statements (TRUE or FALSE statements). Lets start by seeing if anyone has a height greater than 70 inches.

```{r}
data$height_in > 70
```
 
Since there is someone taller that 70 inches, let's see who it is and the rest of the data

```{r}
#to see data for only people with height > 70 inches:
data[data$height_in > 70,] 
#notice that we need to include the comma because we want to see rows with height >70

#to see the name of the person taller than 70 inches:
data[data$height_in > 70,]$names
```

We can combine our criteria for height and age. Say we want to know if someone is either taller 70 inches *or* older than 50 years old (they fit one or the other criterion, but don't have to meet both). Alternatively we could also see if anyone is both taller 70 inches *and* older than 50 years old (they must meet both criteria).

```{r}
data[data$height_in > 70 | data$age > 50,]$names #or
data[data$height_in > 70 & data$age > 50,]$names #and
```

We could also create a new data frame with only people shorter that 70 inches:

```{r}
data_short <- data[data$height_in < 70,] 
```


Now that we have a data frame, we can add to it. Remember how we can perform functions on vectors? The same goes for columns in a data frame. Here's how we might create a column to tell us height in cm:

```{r}
data$height_cm <- 2.54*data$height_in
```


#### Reading in Data

You will be happy to know our main source of data frames is not this cumbersome way of building them in R. We can import them from various file types including csv (comma separated values) files. I generally enter my data into an excel file, save it as a CSV, and read it in using the read.csv function.

In the same Wyo Courses module as this file, you will find a file called "countries_aggregated8.21.20.csv". Download this spreadsheet and save it in the same folder as this R markdown/R project (which should be your working directory). This spreadsheet contains numbers of COVID-19 cases confirmed in each country each day since January 22, 2020. We will use this data to demonstrate some of the things we can do with data frames in R including plotting. For more COVID-19 data check out https://datahub.io/core/covid-19. 

Let's read in the data using the code below. If you need to, feel free to double check where your working directory is (getwd()) as this is a common source of error messages. Note that as long as the data is in your working directory (and not within a folder in that directory), you do not need to write out the entire file path. This is convenient if you want to share code and files because the other person will not need to change a file path every time they want to read in data. 

```{r}
covid_data <- read.csv("countries_aggregated8.21.20.csv")
```

First let's look at the structure of our data frame. What type of data is present in each column? This is important because often data doesn't import the exact was we might want it to. Pay special attention to the "Date" column. You can also view the whole data frame by clicking on the spreadsheet icon by "covid_data" in the environment pane.

```{r}
head (covid_data) #this shows us the first several rows of the data frame
str (covid_data) #this tells you the types of data present in each column
View (covid_data) #this opens a tab to view your data frame
```

Take a look back at our "Hierarchy of data types" you might notice that dates and time are not on this list. Therefore, R treats date or time data as essentially a character string or categorical data. Obviously dates have much more meaning than this (they at least go in a certain order...) so we will use the "lubridate" package you installed earlier to help R interpret date and time data. 

#### *Challenge #x:*
##### *Check to make sure the lubridate package is loaded, if not, load it again*

```{r}
library (lubridate)
```



The following code will essentially tell R to treat the data in this column as a date rather than just a categorical variable. The function ymd() is part of the lubridate package, stands for Year, Month, Day, and denotes the format of the date. 

```{r}
covid_data$Date <- ymd (covid_data$Date)
```


#### *Challenge #x:*
##### *What class of data is now in the "Date" column?*

```{r}
class (covid_data$Date)
```


#### *Challenge #x:*
##### *Can you subset our COVID data frame to only contain data from the United States? (feel free to ask for help and/or reference how we did this to our "Sally, Bob, Joe, Susan" data frame)*

```{r}
covid_data_us <- covid_data[covid_data$Country=="US",]
```


### PLOTTING!!!

This is my favorite part so I hope you're as excited as I am. We will start by using the plot() function in base R to plot cases in the US over time. The most simple way to do this is shown below:

```{r}
plot (Confirmed ~ Date, data=covid_data_us)
```

I don't know what you think, but I think we can do better than that graph. Let's add some color and clean it up a bit.

```{r}
plot (Confirmed ~ Date, data=covid_data_us, type="l", col="darkblue",
      xlab="Confirmed Cases", main="COVID cases over time in the US")
```


#### *Challenge #x:*
##### *Go back to the help page for the plot() function. What does type="l" do? Look at some of the other arguments I added too.*


Lets add some more data to our plot. Once we have an established plot, we can add more data to it using the points() and lines() functions. These take most of the same arguments as the plot() function. 

```{r}
#plotting confirmed cases
plot (Confirmed ~ Date, data=covid_data_us, type="l", col="darkblue",
      ylab="Number of Cases", main=" Total COVID cases over time in the US")
#adding lines for deaths
lines (Deaths ~ Date, data=covid_data_us, col="red")
#adding lines for recovered cases
lines (Recovered ~ Date, data=covid_data_us, col="forestgreen")
#adding a legend
legend ("topleft", legend=c("Confirmed", "Deaths", "Recovered"), col=c("darkblue", "red", "forestgreen"), lty=1)
```

#### *Challenge #x:*
##### *Can you add Recovered cases to the above plot? Try to add it in the legend to!*


### GGPLOT

Up until now we have been using base R to plot. There is a lot we can do with R's built-in plotting functions, but it is somewhat limited. This is why I often use the ggplot2 package for most of the figures I make. The coding is a bit more confusing at first but it's well-worth it for the pretty graphs!! In case you need a reminder: here's my favorite source of amazing ggplot (and base R) graphics ideas and code from the first day of class: https://www.r-graph-gallery.com/ggplot2-package.html


#### *Challenge #x:*
##### *Don't forget to make sure this package is loaded and installed!*

```{r}
library (ggplot2)
```


Let's start off by making a similar plot to the above but with ggplot. In ggplot we begin by defining what we want to plot using the function ggplot(). We tell it the name of our data frame and define our x and y variables as "aesthetics":

```{r}
(us_plot2 <- ggplot (data=covid_data_us, aes (x=Date, y=Confirmed)))
```

You'll notice that this only shows a blank plot (no data). From here we can add components (data points, lines, themes, labels, etc.). Click through the resulting plots to see how your graph changes with each addition.

```{r}
(us_plot2 <- us_plot2 + geom_line (col="blue"))
(us_plot2 <- us_plot2 + theme_classic())
(us_plot2 <- us_plot2 + labs (y="Confirmed Cases", title="COVID cases over time in the US"))
```


#### *Challenge #x:*
##### *Try copying and pasting the code below (it's the exact same as the code you ran in the above two chunks) into the console and running it. Notice where your graphs appear. You can use the blue arrows at the top of the "plots" pane to click through each version of the plot*

```{r eval=FALSE, include=FALSE}
(us_plot2 <- ggplot (data=covid_data_us, aes (x=Date, y=Confirmed)))
(us_plot2 <- us_plot2 + geom_line (col="blue"))
(us_plot2 <- us_plot2 + theme_classic())
(us_plot2 <- us_plot2 + labs (y="Confirmed Cases", title="COVID cases over time in the US"))
```

ggplot makes it very easy to graph way too many things at once. If you wanted to plot every single country in a different color you can simply define this in the aesthetics. Also notice that it is possible to add multiple things at once to you plot rather than re-defining your plot each time (I actually prefer to do it this way).

Warning: this code may take a while to run.

```{r}
world_plot <- ggplot (data=covid_data, aes (x=Date, y=Confirmed, color=Country)) + 
  geom_line() + 
  theme_classic() +
  theme (legend.position = "none") #the legend was way too big and not super helpful...

world_plot
```

That graph is a mess, what if we only wanted to plot countries with more than, say, half a million cases at the most recent date ()? First let's create a subset of the data with only our countries of interest

```{r}
#create a vector of country names
top_countries <- covid_data[covid_data$Date=="2020-08-18" & covid_data$Confirmed>500000,]$Country
#create a data frame that is a subset of covid and only includes our countries of interest
covid_dropped <- covid_data[covid_data$Country %in% top_countries,] 
#using %in% is very similar to the logical arguements we've used earlier, 
#it looks for countries in covid$Countries and keeps only the ones that are 
#present in the top_countries vector
```

```{r}
world_plot2 <- ggplot (data=covid_dropped, aes (x=Date, y=Confirmed, color=Country)) + 
  geom_line() + 
  theme_classic() 
world_plot2
```

One more thing: To demonstrate another plot type we can make in ggplot, let's plot case fatality rates among our top countries! First let's extract the data we need. I'm going to let you do this on your own. We will do this for the most recent data (on Date=="2020-8-18"), and we will need to create a new column for fatality rates. We are defining case fatality rates as the number of deaths divided by the total confirmed cases.


#### *Challenge #x:*
##### *Can you create a subset of our "COVID_dropped" data that only includes the most recent date?*
##### *Can you add do this data a column displaying fatality rates?*
##### *Remember your learning coaches are here to help!!!*

```{r}
#subsetting only the most recent data
covid_dropped_recent <- covid_dropped[covid_dropped$Date=="2020-8-18",]
covid_dropped_recent$Fatality_rates <- covid_dropped_recent$Deaths/covid_dropped_recent$Confirmed
```

```{r}
fatality_plot <- ggplot (data=covid_dropped_recent, aes (x=Country, y=Fatality_rates)) + 
  geom_bar(stat="identity")
  
fatality_plot
```


#### *Challenge #x*
##### *Clean up this graph: consider adding a title, adjusting x and/or y labels, choosing a theme, and playing with colors! Google around to find all the things you can do.*

```{r}
fatality_plot <- ggplot (data=covid_dropped_recent, aes (x=Country, y=Fatality_rates)) + 
  geom_bar(stat="identity", fill="darkblue") +
  labs (y="Case Fatality Rate", title="Case fatality rates in countries with the most cases") +
  theme_classic()
  
fatality_plot
```


### Statistics

While graphing data is an excellent tool to explore and communicate data, statistical analyses are also important to determine whether patterns we see are spurious or might actually mean something.

To demonstrate some basic statistical analyses we can do in R, we will use another COVID-related data set. This new data set includes the number of confirmed cases by county in the United States (as of August 18th). It also includes data from the 2016 presidential election, population density, and mask use in each county. Mask use is from a New York Times survey and is listed as the proportion of people who report "Never", "Rarely", "Sometimes", "Frequently", or "Always" wearing a mask in public when they will come within 6 feet of another person. 


#### *Challenge #x*
##### *Read in the "covid_2016election_data.csv" file. Make sure it's saved in the same directory as everything else we've been working on. To make you're life easier moving forward, name the object "elect"*

```{r}
elect <- read.csv ("covid_2016election_data.csv", header=TRUE)
```

```{r}
elect$Party <- ifelse (elect$votes_dem > elect$votes_gop, "Blue", "Red")
```


#### t-tests and ANOVA

These are analyses we can use with categorical variables. In our example, we have categorized counties as "Blue" or "Red" based on which presidential candidate received the most votes in 2016. We can use a t-test or ANOVA to determine whether the number of COVID cases is different in counties with different political leanings. In this case, we only have two categories, which allows us to use a t-test. If we had more than two categories, or wanted to explore more than one way of categorizing counties, we would need to use an ANOVA. The two analyses should give a similar result here.

Note that we will transform the data of confirmed cases by taking the natural log so that the data is more normally distributed. 

```{r}
#t.test example
t.test (log (Case) ~ Party, data=elect)
```

This tells us that that the average number of cases in "Blue" states is higher than that in "Red" states and the p value  of 1.665e-10 (less than 0.05) tells us that we can consider this difference "statistically significant". We can easily visualize this using a box plot:

```{r}
#base R
boxplot (log (Case) ~ Party, data=elect, ylab="log of Confirmed Cases")

#ggplot
ggplot (elect, aes (x=Party, y=log (Case), color=Party)) + 
  geom_boxplot() + 
  theme_classic() +
  theme (legend.position = "none") +
  scale_color_manual (values = c( "#000099", "#CC0000")) +
  #notice we can specify colors using this 6 character system, google around to find tons of nice colors
  scale_x_discrete(labels=c ("Blue"="Democrat", "Red"="Republican")) +
  labs (y="log of Confirmed Cases", x="Winner of 2016 Presidential Election")
```

For reference, here is some code that you could use to run the analysis using an ANOVA. It should give essentially the same result as above.

```{r}
#anova example:
anova (lm (Case ~ Party, data=elect))
```

#### *Challenge #x*
##### *I encourage you to think about what might cause the difference in case number we demonstrate above. A place to start is thinking about differences in population and population density. Our data set includes population density ("Population_Density"), does this differ significantly between Blue and Red counties? Use a t-test and make a box plot to test this hypothesis. I recommend log transforming this variable as well*

```{r}
#t test
t.test (log (Population_Density) ~ Party, data=elect)

#ggplot
ggplot (elect, aes (x=Party, y=log (Population_Density), color=Party)) + 
  geom_boxplot() + 
  theme_classic() +
  theme (legend.position = "none") +
  scale_color_manual (values = c( "#000099", "#CC0000")) +
  scale_x_discrete(labels=c ("Blue"="Democrat", "Red"="Republican")) +
  labs (y="log of Confirmed Cases", x="Winner of 2016 Presidential Election")
```


#### Linear Models

We have related both confirmed COVID-19 case numbers and population density to political leanings of counties, but how do these two variable relate directly to each other? These are both continuous variables so we can use a linear model. 

We will log-transform both of these variables. 

```{r}
#notice how we can store the model for further use
lm1 <- lm (log (Case) ~ log (Population_Density), data=elect)
#a good way to learn about your linear model is using the summary function:
summary (lm1)
```

Under the "Estimate" column, we get our intercept (Intercept) and slope (log(Population_Density)) that would give the equation of the line that would fit our data. The positive slope tells use that population density and COVID cases are positively associated (Cases increase as population density increases). Again, we have a p-value of way less than 0.5 so we would consider this association "significant"

Let's plot this data and add the line!

```{r}
#base R:
plot (log (Case) ~ log (Population_Density), data=elect, ylab="log of Confirmed Cases", xlab="log of Population Density")
abline (lm1)

#ggplot:
ggplot (elect, aes (x=log (Population_Density), y=log (Case))) + 
  geom_point() +
  geom_smooth (method=lm) +
  labs (x="log of Population Density", y="log of Confirmed Cases") +
  theme_classic ()
```


#### *Challenge #x*
##### *Our data also includes data on mask use in each county. How does this relate to the number of confirmed COVID cases?*
*Disclaimer: technically speaking, we should consider using a slightly different model as this is proportional data but this general linear model with at least give us an idea of what's going on.*


```{r}
#notice how we can store the model for further use
lm2 <- lm (log (Case) ~ log (Frequently), data=elect)
#a good way to learn about your linear model is using the summary function:
summary (lm2)

#ggplot:
ggplot (elect, aes (x=Frequently, y=log (Case))) + 
  geom_point() +
  geom_smooth (method=lm) +
  labs (x="Percent of people 'Frequently' wearing a mask", y="log of Confirmed Cases") +
  theme_classic ()
```

Congratulations! You have completed the tutorial. Go up to the top of this panel in R Studio and click preview. This will give you a preview of you completed R tutorial and resource guide. You will also notice a file with the same name as this R markdown in the same folder with a .html extension. You can open this (it will open in an internet browser) and use it as a nice source of code and information whenever you need. Good luck on your future coding!

If you are interested in a more in-depth tutorial, there are many great ones online including this one: 
https://datacarpentry.org/R-ecology-lesson/index.html
